Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.qmd
Add captions

![CAPTION - FILL THIS IN!](./figures/methionine-reactions.png){#fig-methionine-reactions}

@@

![CAPTION: FILL THIS IN!](./figures/laplace.png){#fig-laplace}

@@

![CAPTION: FILL THIS IN!](./figures/missing.png){#fig-missing}

@@

![CAPTION: FILL THIS IN!](./figures/decomposition.png){#fig-decomposition}

>>>>

::: {#fig-methionine-reactions layout-ncol=1}

![](./figures/methionine-reactions.png)

The methionine cycle as modelled, with the solid black lines representing the
reactions, the green lines representing allosteric interaction, and the red
lines representing allosteric inhibition. The bold fonts are the reaction names
and the regular font represents the metabolites.

:::

@@

::: {#fig-laplace layout-ncol=1}

![](./figures/laplace.png)

CAPTION: FILL THIS IN!

:::

@@

::: {#fig-missing layout-ncol=1}

![](./figures/missing.png)

Results of removing concentration measurements for the metabolite $ahcys_c$
from our case study dataset.  (**Top**) A comparison of metabolite concentration
residuals between the full measurement dataset (blue-green) and the missing-data
dataset (grey), displayed on natural logarithmic scale. The missing-data
model was unable to predict the withheld $ahcys_c$ concentrations.  (**Bottom
Left**) The marginal posterior distribution for the Michaelis constant
$K_m^{AHC1,ahcys_c}$ in each model, alongside the true parameter value used to
generate both datasets. The true value is recovered by the complete-data model
but not by the missing-data model.  (**Bottom Right**) The distribution of total
log-likelihood for out-of-sample flux measurements in both models. There is a
significant overlap between the two distributions, suggesting that removing the
$ahcys_c$ measurement did not cause catastrophic prediction failure. However,
overall the complete-data model tended to make better predictions.

:::

@@

::: {#fig-decomposition layout-ncol=1}

![](./figures/decomposition.png)

Illustration of how analysing a system with Maud can yield actionable insights
about the underlying metabolic network.  (**Top Left**) A schematic of the
regulatory interactions associated with the enzyme $GNMT1$. Dashed green
lines represents allosteric activation, dashed red lines indicate allosteric
inhibition and solid red lines represent competitive inhibition. (**Bottom
Left**) Comparison of marginal posterior distributions for $GNMT$ flux in
conditions 1 and 2. (**Right**) Log-scale ratios of the regulatory elements
defined in equation \eqref{eq-decomposition}. Note that the $Reversibility$
and $k_cat$ components are excluded: this is because this reaction was modelled
as irreversible, and $k_{cat}$ was modelled as constant across conditions.
These plots identify why flux in condition 2 is higher than in condition 1: the
flux increase is due to allostery and saturation with no control from enzyme
concentration changes.

:::

### 2023-10-27 paper/main.tex
Add captions

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

{\centering \includegraphics{./figures/methionine-reactions.png}

@@

\caption{\label{fig-methionine-reactions}CAPTION - FILL THIS IN!}

@@

{\centering \includegraphics{./figures/laplace.png}

@@

@@

{\centering \includegraphics{./figures/missing.png}

@@

\caption{\label{fig-missing}CAPTION: FILL THIS IN!}

@@

{\centering \includegraphics{./figures/decomposition.png}

@@

\caption{\label{fig-decomposition}CAPTION: FILL THIS IN!}

>>>>

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{./figures/methionine-reactions.png}

@@

\end{minipage}%

\caption{\label{fig-methionine-reactions}The methionine cycle as
modelled, with the solid black lines representing the reactions, the
green lines representing allosteric interaction, and the red lines
representing allosteric inhibition. The bold fonts are the reaction
names and the regular font represents the metabolites.}

@@

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{./figures/laplace.png}

@@

\end{minipage}%

@@

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{./figures/missing.png}

@@

\end{minipage}%

\caption{\label{fig-missing}Results of removing concentration
measurements for the metabolite \(ahcys_c\) from our case study dataset.
(\textbf{Top}) A comparison of metabolite concentration residuals
between the full measurement dataset (blue-green) and the missing-data
dataset (grey), displayed on natural logarithmic scale. The missing-data
model was unable to predict the withheld \(ahcys_c\) concentrations.
(\textbf{Bottom Left}) The marginal posterior distribution for the
Michaelis constant \(K_m^{AHC1,ahcys_c}\) in each model, alongside the
true parameter value used to generate both datasets. The true value is
recovered by the complete-data model but not by the missing-data model.
(\textbf{Bottom Right}) The distribution of total log-likelihood for
out-of-sample flux measurements in both models. There is a significant
overlap between the two distributions, suggesting that removing the
\(ahcys_c\) measurement did not cause catastrophic prediction failure.
However, overall the complete-data model tended to make better
predictions.}

@@

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{./figures/decomposition.png}

@@

\end{minipage}%

\caption{\label{fig-decomposition}Illustration of how analysing a system
with Maud can yield actionable insights about the underlying metabolic
network. (\textbf{Top Left}) A schematic of the regulatory interactions
associated with the enzyme \(GNMT1\). Dashed green lines represents
allosteric activation, dashed red lines indicate allosteric inhibition
and solid red lines represent competitive inhibition. (\textbf{Bottom
Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
in conditions 1 and 2. (\textbf{Right}) Log-scale ratios of the
regulatory elements defined in equation \eqref{eq-decomposition}. Note
that the \(Reversibility\) and \(k_cat\) components are excluded: this
is because this reaction was modelled as irreversible, and \(k_{cat}\)
was modelled as constant across conditions. These plots identify why
flux in condition 2 is higher than in condition 1: the flux increase is
due to allostery and saturation with no control from enzyme
concentration changes.}
